---
title: "Reproducible Research Project 1"
author: "J. Hampton"
date: "February 22, 2018"
output: html_document
---

## Load data file, relevant libraries and convert the steps column of the data into numeric.

```{r}
echo=TRUE
setwd("~/RProjects/Coursera/ReproducibleResearch Course5")
ActivityData<-read.csv("activity.csv")
library(dplyr)
library(ggplot2)
ActivityData <- ActivityData %>% mutate(steps = as.numeric(steps))
```

## Q1.  Calculate the total number of steps taken per day, create a histogram and show statistics based on totalnumber of steps per day.  

```{r}
DailySteps <- ActivityData %>% group_by(date) %>% summarise(sum(steps))
ggplot(data=DailySteps, aes(x=`sum(steps)`)) + 
  geom_histogram() + 
  ggtitle("Histogram of daily steps") + 
  labs(x="Daily Steps")
 summary(DailySteps)
```

Median steps per day = 10765
Mean steps per day = 10766
 
 
## Q2. What is the average daily activity pattern? 
 
```{R}
Pattern <- ActivityData %>% na.omit %>% group_by(interval) %>% summarise(average = mean(steps))
Pattern %>% ggplot(aes(x=interval, y=average)) + geom_line() + labs(title = "Average daily activity pattern", x = "time interval(min)", y= "average steps")
Pattern %>% filter(average == max(average))
  
```
The 5-minute interval of 835 minutes contains the maximum number of steps.

## Q3. Imputing missing data

First, the total number of missing values (NA's) in the dataset is calculated.

```{R}
sum(is.na(ActivityData$steps))

```
The total number of missing values:2304

Replacing NA's in the dataset was done by calculating the interval mean using the existing data, then replacing the NA's with those calculated means.

```{R}
imputedAD <- ActivityData %>% group_by(interval) %>% mutate(impute_steps = mean(steps, na.rm=TRUE))
imputedAD$steps[is.na(imputedAD$steps)] <- imputedAD$impute_steps[is.na(imputedAD$steps)]
```

Create dataset with total steps per day with imputed data and create histogram.  Report mean and median total steps per day and compare to mean and median with original dataset.

```{R}
TotalImputed <- imputedAD %>% group_by(date) %>% summarise(allImputed = sum(steps, na.rm=TRUE))
summary(TotalImputed)

ggplot(data=TotalImputed, aes(x=allImputed)) + 
  geom_histogram() + 
  ggtitle("Histogram of daily steps with imputed data") + 
  labs(x="Daily Steps")
```

The imputed data appeared to have a higher impact on the 1st quartile with only a 1 point increase in the median and no change to the mean.

## Q4.  Are there differences in activity patterns between weekdays and weekends?

```{R}
ActivityData$weekday <- ifelse(weekdays(as.Date(ActivityData$date)) %in% c("Saturday","Sunday"),"weekend", "weekday") 

StepsByDay <- ActivityData %>% group_by(interval, weekday) %>% summarize(average = mean(steps, na.rm=TRUE))

ggplot(StepsByDay, aes(x=interval, y=average, colour = weekday)) + geom_line() +
     facet_wrap(~weekday, ncol=1)  +
    labs(x="Time", y = "Average number of steps" )
```


